% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\documentclass[
]{article}
\usepackage{xcolor}
\usepackage{amsmath,amssymb}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\usepackage{tikz}
\usepackage{pgfplots}
\usetikzlibrary{arrows.meta,positioning,shapes.geometric,fit,calc}
\usepgfplotslibrary{groupplots}
\pgfplotsset{compat=1.18}
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\author{}
\date{}

\begin{document}

\section{Problem-Shaped Virtual Machines for Exact Local Transformer
Execution}\label{problem-shaped-virtual-machines-for-exact-local-transformer-execution}

\textbf{Aviraj Khare}\\
Email: \texttt{avirajkhare00@gmail.com}\\
X: \texttt{@avirajkhare00} (\texttt{x.com/avirajkhare00})

\subsection{Abstract}\label{abstract}

Small local transformers are a poor fit for exact tasks when they are
asked either to guess a final answer in one shot or to imitate a full
general-purpose machine. This paper argues for a middle ground: build a
small exact machine for the task first, then use the model only to guide
ambiguous choices inside that machine. We call that machine a
\textbf{problem-shaped virtual machine (PSVM)}. A PSVM keeps only the
steps that matter for one family of tasks. The runtime still checks
correctness, generates valid next steps, and handles rollback. The model
does not replace execution. It helps choose between valid options, for
example by ranking search branches or field candidates. The repository
demonstrates this pattern with browser-local runtimes for Sudoku,
invoice and receipt extraction, Tally-style voucher extraction, and
Weiqi. The main claim is simple: for narrow exact tasks, small models
are more useful as guides inside a task-specific machine than as
free-running solvers over raw inputs or generic machine traces.

\subsection{Paper Type and Scope}\label{paper-type-and-scope}

This document should be read as an \textbf{implementation and systems
paper}. It argues from working code in the repository, not from a
finished benchmark study and not from a universal VM spec. Its main
design claim is:

\begin{itemize}
\tightlist
\item
  build the machine around the task
\item
  keep correctness checks in exact code
\item
  train the model on the ambiguous choices, not on irrelevant machine
  detail
\end{itemize}

The companion document for machine and trace semantics is
\href{/Users/avirajkhare/hack2/transformers/transformer-in-notion/docs/psvm-yellow-paper.md}{PSVM
Runtime and Trace Semantics}. That document is narrower than this paper.
It defines the shared runtime contract, trace format, verifier behavior,
and model/runtime interfaces without claiming that all task families
should share one universal machine.

\subsection{AI Assistance Disclosure}\label{ai-assistance-disclosure}

This paper was drafted and edited with AI assistance. The author
directed the scope, selected the claims, validated the implementation
references against the repository, and is responsible for the final
wording, technical judgments, and positioning.

\subsection{One-Sentence Thesis}\label{one-sentence-thesis}

For narrow exact tasks, the right target for a local model is usually
not the final answer and not a full machine trace. It is the smallest
machine that can still run the task correctly while exposing the
decisions a model can help with.

\subsection{Motivation}\label{motivation}

Exact tasks are common in browser software: puzzle solving, rule
checking, pricing, validation, scheduling, and other deterministic
workflows. These tasks have two properties that matter here.

In this paper, a \textbf{deterministic workflow} just means a task where
rules can check whether a step or final result is correct. Examples
include:

\begin{itemize}
\tightlist
\item
  Sudoku solving
\item
  invoice math
\item
  receipt total extraction
\item
  Tally voucher extraction
\item
  tax calculation
\item
  configuration validation
\item
  approval flows with exact rules
\end{itemize}

First, they have a real notion of correctness. A step is legal or
illegal. A branch is valid or contradictory. The final output either
satisfies the rules or it does not.

Second, they often need far less machinery than a general-purpose
computer. A Sudoku solver does not need arbitrary pointer arithmetic. A
receipt total extractor does not need a large generic bytecode machine.
A Tally voucher extractor does not need free-form text generation for
every field.

The usual formulations miss that structure.

\subsubsection{One-shot prediction is too
coarse}\label{one-shot-prediction-is-too-coarse}

If the model is asked to map directly from task instance to final
answer, all of the exact intermediate transitions are hidden. That makes
supervision weak and error analysis vague. It also makes rollback,
legality checks, and browser trace visualization much harder.

\subsubsection{Full-machine execution is too
broad}\label{full-machine-execution-is-too-broad}

If the model is asked to emulate a full VM, compiler IR, or broad
bytecode surface, it must learn large amounts of machine behavior that
the domain never needs explicitly. Traces become longer, token entropy
rises, and more of the model budget goes into machine scaffolding
instead of task semantics.

The practical question is therefore not:

\texttt{can\ a\ transformer\ execute\ a\ machine?}

The practical question is:

\texttt{what\ is\ the\ smallest\ executable\ machine\ that\ still\ preserves\ this\ task\textquotesingle{}s\ truth\ conditions?}

That machine is what this note calls a \textbf{problem-shaped virtual
machine}.

\begin{figure}[t]
\centering
\resizebox{\linewidth}{!}{%
\begin{tikzpicture}[
  box/.style={draw, rounded corners=6pt, align=left, text width=0.27\linewidth, minimum height=5.6cm, inner sep=8pt, line width=0.8pt},
  chip/.style={draw=none, rounded corners=4pt, font=\bfseries\small, inner xsep=8pt, inner ysep=4pt},
  note/.style={font=\small\itshape},
]
\node[box, fill=red!4] (oneshot) {
  \tikz{\node[chip, fill=red!18] {Guess};}\\[5pt]
  \textbf{One-shot prediction}\\[4pt]
  Input $\rightarrow$ final answer\\[6pt]
  \textit{Strengths}\\
  - simple interface\\
  - low system complexity\\[4pt]
  \textit{Weaknesses}\\
  - hides intermediate state\\
  - hard to verify or debug\\
  - illegal outputs are easy
};
\node[box, fill=green!6, right=1.0cm of oneshot] (psvm) {
  \tikz{\node[chip, fill=green!18!white] {Guide};}\\[5pt]
  \textbf{PSVM middle ground}\\[4pt]
  Input $\rightarrow$ exact state $\rightarrow$ valid options $\rightarrow$ ranked choice\\[6pt]
  \textit{Strengths}\\
  - exact runtime keeps correctness\\
  - model helps only with ambiguity\\
  - trace is inspectable\\[4pt]
  \textit{Weaknesses}\\
  - requires task-specific design\\
  - still needs good candidate coverage
};
\node[box, fill=blue!4, right=1.0cm of psvm] (fullvm) {
  \tikz{\node[chip, fill=blue!18] {Emulate};}\\[5pt]
  \textbf{Full VM emulation}\\[4pt]
  Input $\rightarrow$ broad machine trace\\[6pt]
  \textit{Strengths}\\
  - very general target\\
  - can model full execution\\[4pt]
  \textit{Weaknesses}\\
  - too much irrelevant machine detail\\
  - longer traces\\
  - more context and decoding risk
};
\draw[-{Latex[length=3mm]}, thick] (oneshot.east) -- (psvm.west);
\draw[-{Latex[length=3mm]}, thick] (fullvm.west) -- (psvm.east);
\node[note, above=0.2cm of oneshot.north] {too coarse};
\node[note, above=0.2cm of psvm.north] {right-sized};
\node[note, above=0.2cm of fullvm.north] {too broad};
\end{tikzpicture}
}
\caption{PSVMs sit between one-shot answer prediction and full-machine emulation.}
\label{fig:design-space}
\end{figure}

\subsection{What a Problem-Shaped VM
Is}\label{what-a-problem-shaped-vm-is}

A PSVM is a \textbf{small task-specific machine}. It is built for one
family of tasks and keeps only the steps that matter for that family. It
is not general on purpose.

A PSVM usually has five parts:

\begin{itemize}
\tightlist
\item
  \textbf{state}: the minimum information needed to continue the task
\item
  \textbf{valid next steps}: the actions that make sense from the
  current state
\item
  \textbf{verifier}: exact code that checks whether a step is valid
\item
  \textbf{trace}: a fixed log of states and decisions
\item
  \textbf{model hook}: the place where a model helps choose between
  valid options
\end{itemize}

For this repository, the useful design rule is:

\texttt{trim\ the\ machine\ to\ the\ problem}

That means:

\begin{itemize}
\tightlist
\item
  keep only task steps that carry real meaning
\item
  remove generic machine bookkeeping that the task does not care about
\item
  keep correctness checks in exact code
\item
  make failure and rollback explicit when search is real
\item
  keep the trace format fixed so training and debugging stay stable
\item
  let the model help only where there is real ambiguity
\end{itemize}

Three simple examples make this concrete:

\begin{itemize}
\tightlist
\item
  \textbf{Sudoku}: the state is the board and candidate information; the
  valid next steps are legal placements, branches, and backtracks.
\item
  \textbf{Receipt total extraction}: the state is the OCR text plus
  money candidates; the valid next steps are candidate selections such
  as ``choose this amount as the total.''
\item
  \textbf{Tally voucher extraction}: the state is OCR or layout text
  plus the likely voucher family and legal field candidates; the valid
  next steps are field selections that fit that voucher schema.
\end{itemize}

The machine should be as small as possible, but not smaller than the
task. If a step is required for correctness, keep it. If it exists only
because a broad VM needs it, it is probably noise for this setup.

\subsubsection{What we mean by a trace}\label{what-we-mean-by-a-trace}

A \textbf{trace} is a step-by-step log of execution.

At each step it records enough information to understand what the
machine saw and what it did. Depending on the task, that can include:

\begin{itemize}
\tightlist
\item
  the current state
\item
  the valid options
\item
  the selected option
\item
  the resulting state
\end{itemize}

We keep the trace format fixed so it is useful for training, debugging,
and UI rendering.

\subsection{Why This Fits Transformer
Execution}\label{why-this-fits-transformer-execution}

This note is motivated by a systems intuition rather than a formal
theorem.

An \textbf{autoregressive model} predicts the next token from the tokens
that are already present. That makes it naturally better at extending a
sequence than at behaving like a normal program that can overwrite
arbitrary memory cells.

A conventional interpreter evolves state roughly as:

\texttt{state\ +\ instruction\ -\textgreater{}\ new\ state}

A transformer behaves more like:

\texttt{trace\_t\ -\textgreater{}\ trace\_t+1}

Here:

\begin{itemize}
\tightlist
\item
  \texttt{trace\_t} means the execution log up to step \texttt{t}
\item
  \texttt{trace\_t+1} means the same log with one more record appended
\end{itemize}

The important point is that earlier tokens stay fixed. The model reads
the existing sequence and predicts what comes next. It does not directly
edit memory the way a normal mutable RAM machine does.

That gives a useful mental model:

\begin{itemize}
\tightlist
\item
  tokens can represent state records or decisions
\item
  attention can retrieve the earlier records that matter
\item
  the trace becomes the model's working context
\end{itemize}

This is a design heuristic, not a proof. But it has a concrete
consequence: the trace format and instruction set matter a lot. If the
trace is full of irrelevant machine detail, the model wastes capacity
reading and predicting boilerplate instead of task structure.

That is why a PSVM can be a better fit than either one-shot prediction
or full machine traces.

\subsection{System Pattern}\label{system-pattern}

The intended execution loop is:

\texttt{task\ instance\ -\textgreater{}\ exact\ state\ -\textgreater{}\ valid\ options\ -\textgreater{}\ model\ scores\ options\ -\textgreater{}\ exact\ runtime\ verifies\ and\ applies\ -\textgreater{}\ new\ state\ -\textgreater{}\ trace\ grows}

\begin{figure}[t]
\centering
\resizebox{\linewidth}{!}{%
\begin{tikzpicture}[
  nodebox/.style={draw, rounded corners=6pt, align=center, text width=2.45cm, minimum height=1.15cm, fill=gray!6, line width=0.8pt},
  exact/.style={nodebox, fill=blue!6},
  model/.style={nodebox, fill=green!8},
  tracebox/.style={draw, rounded corners=4pt, align=center, minimum width=2.3cm, minimum height=0.75cm, fill=orange!8, line width=0.7pt},
  lane/.style={draw, rounded corners=8pt, inner sep=7pt, line width=0.9pt},
]
\node[nodebox] (task) {Task\\instance};
\node[exact, right=0.8cm of task] (state) {Exact\\state};
\node[exact, right=0.8cm of state] (valid) {Valid\\options};
\node[model, right=0.8cm of valid] (score) {Model scores\\options};
\node[exact, right=0.8cm of score] (apply) {Runtime verifies\\and applies};
\node[nodebox, right=0.8cm of apply] (newstate) {New state\\+ trace};
\node[tracebox, below=1.45cm of state] (trace1) {$S_t$};
\node[tracebox, right=0.35cm of trace1] (trace2) {$V(S_t)$};
\node[tracebox, right=0.35cm of trace2] (trace3) {$a_t^*$};
\node[tracebox, right=0.35cm of trace3] (trace4) {$S_{t+1}$};
\node[font=\bfseries\small, above=0.12cm of trace2.north] {trace grows by one record};
\node[lane, fit=(state) (valid) (apply), fill=blue!3, draw=blue!55] (exactlane) {};
\node[lane, fit=(score), fill=green!4, draw=green!45!black] (modellane) {};
\node[font=\bfseries\small, above=0.1cm of exactlane.north] {Exact runtime};
\node[font=\bfseries\small, above=0.1cm of modellane.north] {Learned guide};
\draw[-{Latex[length=3mm]}, thick] (task) -- (state);
\draw[-{Latex[length=3mm]}, thick] (state) -- (valid);
\draw[-{Latex[length=3mm]}, thick] (valid) -- (score);
\draw[-{Latex[length=3mm]}, thick] (score) -- (apply);
\draw[-{Latex[length=3mm]}, thick] (apply) -- (newstate);
\draw[-{Latex[length=3mm]}, thick, bend left=24] (newstate.north) to node[above, font=\small] {repeat} (state.north);
\draw[-{Latex[length=2.5mm]}, thick, orange!70!black] (state.south) -- (trace1.north);
\draw[-{Latex[length=2.5mm]}, thick, orange!70!black] (valid.south) -- (trace2.north);
\draw[-{Latex[length=2.5mm]}, thick, orange!70!black] (score.south) -- (trace3.north);
\draw[-{Latex[length=2.5mm]}, thick, orange!70!black] (newstate.south) -- (trace4.north);
\end{tikzpicture}
}
\caption{The PSVM execution loop. Code owns state and verification; the model only ranks valid options.}
\label{fig:psvm-loop}
\end{figure}

The runtime remains in charge throughout.

The runtime still owns:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  state updates
\item
  valid option generation
\item
  correctness checks
\item
  rollback or backtracking
\end{enumerate}

The model owns only:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  ranking
\item
  value estimation
\item
  ambiguity resolution
\end{enumerate}

In the strongest form:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  an exact teacher runtime defines legal states and steps
\item
  the teacher emits fixed state and decision records
\item
  a local model learns to score ambiguous states or options
\item
  the runtime verifies or rejects student proposals
\item
  the UI streams the trace and state changes in real time
\end{enumerate}

This is not ``weights instead of code.'' It is a hybrid design in which
code keeps correctness and the model helps choose between valid options.

When the model is uncertain, the runtime should still be able to fall
back to a deterministic order or a small beam. The model is a guide, not
the source of truth.

\subsubsection{Minimal transition
function}\label{minimal-transition-function}

The basic PSVM loop can be written in a small amount of notation.

Let:

\begin{itemize}
\tightlist
\item
  \(S_t\) be the exact machine state at step \(t\)
\item
  \(V(S_t)\) be the set of valid next actions from that state
\item
  \(m_\theta(S_t, a)\) be the model score for valid action \(a\)
\end{itemize}

Then the simplest guided choice is:

\[
a_t^* = \arg\max_{a \in V(S_t)} m_\theta(S_t, a)
\]

and the exact runtime applies the state transition

\[
S_{t+1} = \delta(S_t, a_t^*)
\]

This captures the main point of the paper:

\begin{itemize}
\tightlist
\item
  the runtime defines the valid action set \(V(S_t)\)
\item
  the model only ranks actions inside that valid set
\item
  the exact transition function \(\delta\) still updates the machine
  state
\end{itemize}

For search tasks, the action set may include branch, fail, undo, or
rollback steps. For extraction tasks, it may include candidate-selection
steps. The notation stays the same even when the domain changes.

\subsection{PSVM Design from First
Principles}\label{psvm-design-from-first-principles}

The clean way to design a PSVM is to work from first principles.

\subsubsection{Start from correctness}\label{start-from-correctness}

Before designing the machine, say what makes an answer correct. If you
cannot write the correctness check clearly, the PSVM is not ready.

\subsubsection{Define the smallest state that can still run the
task}\label{define-the-smallest-state-that-can-still-run-the-task}

Keep only the information the task actually needs to continue. Extra
machine state that exists only for generality is usually noise.

\subsubsection{Define the valid next
steps}\label{define-the-valid-next-steps}

Each step should correspond to a meaningful domain action. If the task
can backtrack, undo, or fail, those transitions should be explicit too.

\subsubsection{Keep the verifier exact}\label{keep-the-verifier-exact}

The runtime should still decide whether a step is valid. Arithmetic,
schema checks, contradictions, and hard constraints belong in code.

\subsubsection{Use a fixed trace format}\label{use-a-fixed-trace-format}

If the same task state can be written multiple ways, training gets noisy
and debugging gets harder. Tie-breaking, candidate ordering, and record
layout should be stable.

\subsubsection{Give the model only the ambiguous
choices}\label{give-the-model-only-the-ambiguous-choices}

The model should spend its capacity on branch ordering, candidate
ranking, or value estimation. It should not have to relearn exact rules
that code can check perfectly.

\subsubsection{Keep search explicit when search is
real}\label{keep-search-explicit-when-search-is-real}

If the task genuinely branches and backtracks, make that visible. Do not
hide real search inside one opaque ``solve'' action.

\subsubsection{Keep examples honest}\label{keep-examples-honest}

Each example in the repo should answer the same questions:

\begin{itemize}
\tightlist
\item
  what is the state?
\item
  what are the valid next steps?
\item
  what checks correctness?
\item
  what does the model decide?
\item
  what does the trace record?
\end{itemize}

That keeps the examples comparable and stops them from turning into
blind demos.

\subsection{What This Repository Already
Demonstrates}\label{what-this-repository-already-demonstrates}

This repository is strongest as a systems note because it already
contains multiple concrete pieces of the PSVM stack.

\subsubsection{Sudoku}\label{sudoku}

The Sudoku example shows the cleanest version of the argument:
\textbf{model-guided exact search}. The runtime still owns candidate
generation, legality, contradictions, and backtracking. The model only
helps choose which branch to try first.

That is not a small difference. On the current hard preset
\texttt{Forum\ hardest\ 1106\ ·\ r365}, the shipped imitation
transformer required 7,523 ranked branch decisions and 86,376 guided
backtracks, while the regret-trained transformer required 5,133 branch
decisions and 57,057 backtracks on the same exact runtime, cutting wall
time from 96.99 seconds to 64.94 seconds. The rules did not change. The
only thing that moved was branch ordering.

That is exactly the behavior this paper wants from the learned
component: not free-running next-op generation, but better value
estimates over exact PSVM states.

\begin{figure}[t]
\centering
\begin{tikzpicture}
\begin{groupplot}[
  group style={group size=3 by 1, horizontal sep=1.4cm},
  width=0.27\textwidth,
  height=5.2cm,
  ybar,
  ymin=0,
  enlarge x limits=0.35,
  symbolic x coords={Imitation,Regret},
  xtick=data,
  xticklabel style={font=\small, rotate=20, anchor=east},
  nodes near coords,
  every node near coord/.append style={font=\scriptsize},
  ylabel style={font=\small},
  title style={font=\small\bfseries},
]
\nextgroupplot[title={Branch decisions}, ylabel={count}]
\addplot[fill=blue!45] coordinates {(Imitation,7523) (Regret,5133)};
\nextgroupplot[title={Backtracks}, ylabel={count}]
\addplot[fill=orange!65] coordinates {(Imitation,86376) (Regret,57057)};
\nextgroupplot[title={Wall time}, ylabel={seconds}]
\addplot[fill=green!55!black] coordinates {(Imitation,96.99) (Regret,64.94)};
\end{groupplot}
\end{tikzpicture}
\caption{Sudoku benchmark on the hard preset \texttt{Forum hardest 1106 · r365}. Only the learned branch-ordering policy changes.}
\label{fig:sudoku-benchmark}
\end{figure}

\subsubsection{Invoice calculation}\label{invoice-calculation}

The invoice and receipt lanes show that the same pattern works outside
puzzles. Here the runtime first builds legal money candidates from OCR
or structured text, and the model helps choose between those candidates.
The model is not asked to invent totals from scratch.

This matters because it shows that PSVMs are not only about search
problems. They also fit narrow business workflows where exact
arithmetic, exact field legality, and explicit candidate surfaces matter
more than free-form generation.

\subsubsection{Tally voucher extraction}\label{tally-voucher-extraction}

The \texttt{tally/} directory extends the same pattern from one selected
value to a full voucher-shaped record. The runtime first classifies the
voucher family, then builds legal field candidates for that family, and
finally applies a small resolver to keep the chosen fields globally
consistent.

This matters because it shows the PSVM idea in a more realistic document
setting. The model is not asked to generate free-form accounting JSON.
It only ranks candidates inside a runtime that still owns schema checks,
GST logic, field legality, and final record emission.

It also exposes a useful engineering lesson: in this lane, parser
quality and candidate recall matter more than model size. The
adversarial harness makes that visible by separating candidate-missing
failures from ranking failures and resolver failures.

\begin{figure}[t]
\centering
\resizebox{\linewidth}{!}{%
\begin{tikzpicture}[
  nodebox/.style={draw, rounded corners=6pt, align=center, text width=2.1cm, minimum height=1.05cm, fill=gray!6, line width=0.8pt},
  exact/.style={nodebox, fill=blue!6},
  model/.style={nodebox, fill=green!8},
  lane/.style={draw, rounded corners=8pt, inner sep=7pt, line width=0.9pt},
]
\node[nodebox] (ocr) {OCR or\\layout text};
\node[exact, right=0.45cm of ocr] (family) {Voucher\\family};
\node[exact, right=0.45cm of family] (schema) {Schema\\selection};
\node[exact, right=0.45cm of schema] (candidates) {Legal field\\candidates};
\node[model, right=0.45cm of candidates] (ranker) {Local ranker};
\node[exact, right=0.45cm of ranker] (resolver) {Resolver};
\node[exact, right=0.45cm of resolver] (record) {Tally-shaped\\record};
\node[lane, fit=(family) (schema) (candidates) (resolver) (record), fill=blue!3, draw=blue!55] (exactlane) {};
\node[lane, fit=(ranker), fill=green!4, draw=green!45!black] (modellane) {};
\node[font=\bfseries\small, above=0.1cm of exactlane.north] {Exact extraction and validation};
\node[font=\bfseries\small, above=0.1cm of modellane.north] {Learned ranking};
\draw[-{Latex[length=3mm]}, thick] (ocr) -- (family);
\draw[-{Latex[length=3mm]}, thick] (family) -- (schema);
\draw[-{Latex[length=3mm]}, thick] (schema) -- (candidates);
\draw[-{Latex[length=3mm]}, thick] (candidates) -- (ranker);
\draw[-{Latex[length=3mm]}, thick] (ranker) -- (resolver);
\draw[-{Latex[length=3mm]}, thick] (resolver) -- (record);
\end{tikzpicture}
}
\caption{Tally PSVM pipeline. The model ranks candidates; exact code keeps schema checks and final record emission.}
\label{fig:tally-pipeline}
\end{figure}

\subsubsection{Other browser-local
demos}\label{other-browser-local-demos}

The broader repo also includes exact browser-local task demos such as
the Weiqi prototype in \texttt{weiqi/}. Even when those demos do not yet
have a full student training path, they support the same systems
intuition: narrow exact tasks are best expressed as narrow exact
runtimes first.

\subsection{What Is Supported Today, and What Is
Not}\label{what-is-supported-today-and-what-is-not}

This distinction matters if the note is going to be shared outside the
repo.

\subsubsection{Supported now}\label{supported-now}

The codebase already supports these claims:

\begin{itemize}
\tightlist
\item
  browser-local exact runtimes for narrow tasks are practical
\item
  canonical task-shaped traces can be emitted and streamed in the UI
\item
  local model workers can be integrated into that execution loop
\item
  a narrow task-specific instruction surface is implementable across
  more than one domain
\item
  hybrid execution, where the model helps but the runtime remains
  authoritative, is a real engineering pattern rather than only an idea
\item
  a regret-trained value model can sometimes materially reduce exact
  Sudoku search work relative to an imitation-trained transformer on the
  same runtime
\end{itemize}

\subsubsection{Not supported yet}\label{not-supported-yet}

The codebase does \textbf{not} yet justify the stronger claims below:

\begin{itemize}
\tightlist
\item
  that PSVMs already outperform strong baselines quantitatively
\item
  that a student model can replace the teacher runtime end to end
\item
  that the current learned policy wins consistently across puzzle
  regimes
\item
  that the approach scales cleanly from narrow demos to broad arbitrary
  program execution
\item
  that ``compile arbitrary C into weights'' is the immediate right
  target for this stack
\end{itemize}

This is important because the project becomes much more credible when it
says precisely what is and is not already proven.

\subsection{The Paper's Sharpest Claim}\label{the-papers-sharpest-claim}

The most defensible version of the idea is not:

\begin{itemize}
\tightlist
\item
  ``transformers can already execute arbitrary code exactly''
\item
  ``local models can replace exact runtimes''
\item
  ``full VM traces are the right universal target''
\end{itemize}

The sharp claim is:

\textbf{For narrow exact tasks, there is a better intermediate
representation than both direct answer prediction and full-machine
emulation: a problem-shaped VM whose instruction surface already matches
the task's true legal transitions.}

In the strongest form, the learned component over that VM should
estimate state or branch value over exact PSVM states, not merely
imitate the next serialized op.

That claim is strong enough to be interesting and narrow enough to
defend.

\subsection{Experiments That Would Turn This Note into a
Paper}\label{experiments-that-would-turn-this-note-into-a-paper}

To move from a strong systems note to a publishable empirical paper, the
repo needs controlled comparisons.

\subsubsection{Formulations to compare}\label{formulations-to-compare}

For a domain such as Sudoku, compare:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \texttt{state\ -\textgreater{}\ final\ answer}
\item
  \texttt{state\ -\textgreater{}\ next\ domain\ action}
\item
  \texttt{state\ -\textgreater{}\ value/ranking\ over\ legal\ PSVM\ branches}
\item
  \texttt{state\ -\textgreater{}\ next\ token\ of\ a\ broader\ generic\ VM}
\end{enumerate}

The same structure can be repeated for invoice-style deterministic
tasks.

\subsubsection{Metrics}\label{metrics}

The evaluation should emphasize exactness, not only loss:

\begin{itemize}
\tightlist
\item
  exact final solve rate
\item
  average backtracks or downstream search cost
\item
  solve rate under a fixed compute budget
\item
  illegal action rate
\item
  contradiction or verifier-failure rate
\item
  average trace length
\item
  browser latency
\item
  model size versus exactness tradeoff
\end{itemize}

\subsubsection{Ablations}\label{ablations}

The most informative ablations are likely:

\begin{itemize}
\tightlist
\item
  instruction-set size
\item
  trace canonicalization rules
\item
  with versus without exact runtime verification
\item
  op-only versus op-plus-argument prediction
\item
  browser-local inference cost at fixed quality
\end{itemize}

If those comparisons are strong, the paper moves from an interesting
thesis to a measured result.

\subsection{Why This Could Matter}\label{why-this-could-matter}

There is a real gap between two current habits:

\begin{itemize}
\tightlist
\item
  asking models for final answers and hoping they are right
\item
  routing all exact work to external tools or full symbolic engines
\end{itemize}

PSVMs propose a middle layer. They keep symbolic truth in code while
giving the model a narrow, learnable execution surface. That is
especially relevant when the deployment target is the browser, where
latency, memory, and inspectability matter more than broad generality.

The argument is not that every exact task should become a PSVM. The
argument is that many narrow exact tasks already have a small semantic
core, and the model should be trained on that core rather than on
irrelevant machine detail.

\subsection{Honest Positioning If Shared
Now}\label{honest-positioning-if-shared-now}

This note is worth sharing \textbf{now} if it is framed correctly.

The right frame is:

\begin{itemize}
\tightlist
\item
  a research note
\item
  a systems essay
\item
  a workshop or position-paper candidate
\item
  a repository-backed design argument with concrete prototypes
\end{itemize}

The wrong frame is:

\begin{itemize}
\tightlist
\item
  a finished empirical paper
\item
  a claim that arbitrary-code execution has been solved
\item
  a claim that the model already replaces the exact runtime
\end{itemize}

In other words, the idea is shareable today because the conceptual
stance is clear and the repository already contains meaningful evidence.
But the strongest publication target right now is something like a
workshop note, demo paper, or position paper. A full paper needs the
baseline study and quantitative evaluation described above.

\subsection{Conclusion}\label{conclusion}

The repo's most interesting idea is not ``put a transformer in the
browser.'' It is this:

\texttt{trim\ the\ machine\ to\ the\ problem}

If the task is exact and narrow, the model should not be asked either to
guess the final answer in one shot or to emulate a full machine whose
semantics dwarf the problem. It should be asked to operate inside the
smallest executable substrate that still preserves the task's truth
conditions.

That is the case for problem-shaped virtual machines.

\end{document}
